---
output: 
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: '~/Desktop/INFO498Final/paper/svm-latex-ms.tex'
title: "Assesing Health Care Coverage and Access Utilizing the National Health Interview Survey"
author:
- name: John Diego
  affiliation: University of Washington
- name: Warren Wakuzawa
  affiliation: University of Washington
- name:  Adrian Santiago
  affiliation: University of Washington
abstract: "The National Health Interview Survey (NHIS) is the nation’s largest in-person household health survey. It has been conducted annually since 1957 by the National Center for Health Statistics (NCHS), which is a part of the Centers for Disease Control and Prevention (CDC). A broad range of topics are covered but we will specifically focus on healthcare access and expenditure across a multitude of factors such as employer information, race, and academic background."
keywords: "healthcare, NHIS, ACA, medicaid, medicare"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 10pt
# spacing: double
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 9.5)
source('calculations.R')

library(ggplot2)
```

# INTRODUCTION
The Patient Protection and Affordable Care Act (PPACA), commonly shortened to the Affordable Care Act (ACA) and nicknamed Obamacare, is one of the most important healthcare legislature, creating a significant impact on the US Health care system. However, with the arrival of President Trump in the Oval Office, and with the ACA repeal now underway, some speculate huge consequences such as [increased death rates](https://www.washingtonpost.com/posteverything/wp/2017/01/23/repealing-the-affordable-care-act-will-kill-more-than-43000-people-annually/?utm_term=.a2fbb24cd075). In light of these recent events, it is important that we take a look at the current state of our healthcare system and question whether it is failing or not.

The purpose of this paper is to assess healthcare coverage and access across a  multitude of factors. We will specifically look at three aspects of health care - the first aspect is the relationship between educational attainment and access to medical insurance, the second is an overall look at the population and health care coverage, and the final exploration covers healthcare expenditure over the years.

# METHODOLOGY
To answer the questions we layed out, we use data provided from the NHIS survey. We used various subsets of the survey data, depending on the specific question being analyzed.

## **Educational Attainment Vs Healthcare Access**

To answer the question, _"Does higher educational attainment cause greater chance of acquiring medical insurance?"_, we filtered the dataset, only looking a three specific variables, the _highest educational attainment_, _health insurance coeverage_, and _health insurance offered at workplace_. 

Because the dataset was so large, it was difficult to manage and processing the data was time-consuming.  In order to fully combat this, we used the dplyr package in order to modify the dataset we would be using in such a way that only incorporated the columns we needed to address the question at hand.  After, we then filtered the data so that we would have only the information we needed to answer the questions, while weeding out “empty” data such as “not ascertained,” “didn’t know,” as well as “refused.”  When we were certain of the dataset we needed that only contained relevant information to our questions, we then utilized the ggplot2 R package in order to create well-labeled graphs that visually summarized our results.  From the graphs, we were able to draw reasonable conclusions and generate an analysis of why the data was shaped in the particular way it was.  

## **Health Care Access**

In respects to the National Health Interview Survey, we wanted to research what percentage of the population had healthcare coverage. We used the “NOTCOV” column of the survey to retrieve the amount of individuals that had coverage and those who did not have healthcare coverage. We do want to mention that there was a total of 1102 individual that responded with “Don’t know”. WE made the decision to exclude these individuals from our data since they do not represent your targeted population.
 
When comparing just the proportions of those who had healthcare coverage to those who lacked coverage, there was a clear transformation. Only 10.23% of the individuals that participated in this survey did not have healthcare coverage. While 89.76% of these individuals did have coverage. 

## **Health Care Expedentiture**

To gain a general outlook on the healthcare system, one way of looking at it is through the perspective of expenditures. Specifically the question we wanted to explore is, _"How has the amount of money spent for medical care changed over the years?"_ One specific variable that was counted for from the survey was _"Amount family spent for medical care, past 12 months"_. To see this distribution, data from the 2005 to 2015 NHIS surveys was used (N=1,033,155). We then filtered the data further down to those who responded back to the question (N = 994,797). The distribution of health expenditure over the years can be seen in the graph below.

This independent variable only provides a rough estimate on the family's expenditure on medical care, but is still relevant since interviewers directed respondents to exclude the insurance premiums, over-the-counter drugs, and any costs for which they expected to be reimbursed. 

To make calculations on the greater population, we used the calculated weight and strata within the provided by those who constructed the NHIS dataset. According to the [codebook](https://ihis.ipums.org/), the _sample weight_ was calculated using "adjustments for age, race/ethnicity, and sex using the Census Bureau's population control totals", while _strata_ was represents "the impact of the sample design stratification on the estimates of variance and standard errors." A **stratified random sampling** was implemented to calculate the weighted proportions of people who spend X amount for health insurance, X being multiple levels of range of money spent.

# RESULTS

## **Plots** 

```{r expenddistribution, echo=FALSE}
# Line chart
# Change of expend over years (Not weighted)
ggplot(spend.years, aes(x=YEAR, y=count/1000, colour=HCSPENDY)) + geom_line() +
  geom_point(size=.7) +
  labs(x="Year", y="Sum of People (Thousands)", title = "Distribution of Health Expenditure") +
  theme(legend.title=element_blank())
```
**Figure 1. Expenditure Distribution over Time**

\smallskip

```{r weightedexpendproportion, echo=FALSE, fig.height=6}
ggplot(hi.spend, aes(y=factor(YEAR))) +
  geom_point(data= hi.spend, aes(x=zero/population * 100, colour='red'), size=.9) +
  geom_errorbarh(data= hi.spend, aes(x=zero/population * 100, xmin=ci_u.zero/population * 100, 
                                     xmax=ci_l.zero/population * 100),height=0.1) +
  geom_point(data= hi.spend, aes(x=lessthan500/population * 100, colour='gold'), size=.9) +
  geom_errorbarh(data= hi.spend, aes(x=lessthan500/population * 100, xmin=ci_l.lessthan500/population * 100, 
                                     xmax=ci_u.lessthan500/population * 100),height=0.1) +
  geom_point(data= hi.spend, aes(x=spend3000to4999/population * 100, colour='blue'), size=.9) +
  geom_errorbarh(data= hi.spend, aes(x=spend3000to4999/population * 100, xmin=ci_l.spend3000to4999/population * 100, 
                                     xmax=ci_u.spend3000to4999/population * 100),height=0.1) +
  coord_flip() + 
  scale_color_discrete(name="Expenditure", labels = c("$3000 to $4999", "Less than $500", "Zero")) +
  labs(x="Weighted Proportion of People (%)", y="Year", title="Weighted Sum Proportion of Health Care Expenditure")
```
**Figure 2. Expenditure Distribution over Time (Weighted)**

\smallskip

```{r healthcovereduattainment, echo=FALSE}
ggplot(d2015, aes(EDUC1)) + geom_histogram(stat='count') + facet_wrap(~NOTCOV) + 
  ggtitle('Education Level vs. Health Coverage') + labs(x='Education Level') 
```
**Figure 3. Distribution of Health Coverage per Education Level**

\smallskip

```{r workhealthinsuranceeduattainment, echo=FALSE}
ggplot(d2015, aes(EDUC1)) + geom_histogram(stat='count') + facet_wrap(~HIEMPOF) +
  ggtitle('Education Level vs. Workplace-offered Health Insurance') + labs(x='Education Level')
```
**Figure 4. Educational Attainment vs. Workplace-offered Health Insurance**

```{r hi2015, echo=FALSE}
barplot(h, names.arg = m, xlab = "Healthcare Coverage for 2015", 
        ylab="Percentage (%)", col='blue', main = "Percentage of the Respondents that Has Healthcare Coverage")
```

**Figure 5. Percentage of the Respondents that Has Healthcare Coverage**

```{r hi2015employer, echo=FALSE}
barplot(h2, names.arg = m2, xlab = "Healthcare Coverage for 2015", 
        ylab="Percentage (%)", col='blue', main = "Percentage of the Respondents that Has Healthcare Coverage")
```
**Figure 6. Percentage of the Respondents that Has Healthcare Coverage**

# DISCUSSION

## **Educational Attainment Vs Healthcare Access**

The first graph that we see here represents whether or not an individual obtained health coverage based on their education level.  Immediately, we see that the number 2 subgraph contains a lot more results than the number 1 subgraph, which indicates that there were more people who received health coverage than did not. Under the number 1 subgraph, we see the varying number of people who did not receive health coverage given their level of education.  There appears to be a notable peak at the number 14 which is indicative of those who have completed up to a high school diploma.  A second notable peak is at the number 15 which represents those who have attended college but had not received a degree.  Such might be the case because most people finish school up until high school and also attend college, so it would make sense that the bulk of the sample surveyed would reside in that educational attainment categories.  It would furthermore make sense that they did not receive health coverage because nowadays, people obtain health/medical insurance from their workplaces since getting these plans on an individual basis is a lot more expensive.  Those who did not finish college or only obtained a high school diploma might have not been able to get the best occupation due to their lack of education and skills in the workforce, which affected their ability to get health coverage from their employer.  

In the number 2 subgraph, we see three notable peaks--one at the x-coordinate values 14, 15, and 18.  These peaks correspond to the bulk of those who received health/medical insurance given that they either graduated high school, went to college but did not get a degree, or got a Bachelor’s degree.  What is interesting about this is that there were a lot of people who did not AND did get health coverage given that they graduated from high school or went to college but did not get a degree.  This might reveal to us that even with a high school degree it is possible to get a steady job that includes medical/health insurance coverage.  However, analyzing the data further, it makes sense that a lot of people who received health coverage received at least a Bachelor’s degree, because that level of education makes it a lot easier to get jobs that pay well and include a lot of benefits such as medical insurance.

The second graph we are looking at compares educational attainment with whether or not a person received health insurance from their workplace.  The number 1 subgraph represents the count of people who did _receive health insurance at their workplace_ given their highest level of educational attainment, while number 2 subgraph represents the count of people who _did not receive health insurance at work_ given their highest level of educational attainment.  Again, in both graphs, x-coordinate values 14, 15, and 18, corresponding to having graduated high school, attending college without getting a degree, and getting a bachelor’s degree respectively, are the most prominent peaks in the graph.  This is due to the fact that most people finish at least up until high school, attend college at some point and realize that it wasn’t for them, or obtain their bachelor’s degree.  Very _few_ people quit grade school in the middle of it and very few people get a higher education such as a master’s or doctoral degree, which is why the sample count is so low at those other values.  Based on the number 1 subgraph however, we see that the most people who got health insurance from their workplace are those who graduated with a Bachelor’s degree. This reveals the importance of obtaining a Bachelor’s degree when getting a job.  Other interesting points of analysis are that more high school graduates received workplace-offered health insurance than those who attended college but did not get a degree.  One reasoning behind this is that those who did not receive a degree from college might have been those who quit in order to pursue their own businesses, and therefore did not work for an existing corporation that would have otherwise offered them health insurance.  The number 2 subgraph shows that high school graduates were the least likely to receive health insurance from their workplace, which again points back to the fact that they probably weren’t able to get a great enough job that offered them benefits such as health insurance.  Most decent jobs nowadays require people to have at least a bachelor’s degree.  The lowest of the three peaks, though, are those who got a bachelor’s degree, and it shows that with a higher level of education, you will have a _better_ chance of getting a job with better benefits.

ODDS RATIO: 1.19 / 3.66 = **0.33**
The odds of getting coverage being a high school grad is 0.33 times higher than the odds of getting coverage having a bachelor’s degree.

## **Health Care Access**
A huge discussions these recent weeks has been Obama care. Some view it was a total disaster while others suggest that it works but it is not perfect. Our president has spoken that this plan is terrible and this inspired us to consider the what a reality would be like if there was no more Obama care. As a group we believed that without government help, individuals would have to resort to other sorts of health care plans. We decided to focus on employers to study if there is any significant correlation between employers and health care coverage. We do have to note that this information lacked the participation from 42728 individuals. This number was subtracted from the group that did responded as we were interested in the population that did have some sort of coverage. 

As our prior graph indicates, about 89.76% of the surveyed population expressed some sort of healthcare coverage. Out of this population we calculated that 78.69% of population that has some sort of healthcare coverage attains it through an employer. This is only about a 11.07% difference suggesting that the way more people get their healthcare coverage through their workforce. However we did not count those individuals who obtained coverage through their own self-employed business, a union, or even those who were uncertain about their response. 

After analyzing these findings, it is our suggestion that there is some relationship between employers and a person's’ health care coverage. But must point out that it is not very clear to understand how strong this correlation is or to even identify in what specific ways these aspects have an influence on each other. The 11% difference does portray the idea that everyone is obtaining healthcare coverage from their employers. But it is crucial to realize that the portion that missed this question represents, in quantity,  consists of more than half the population that did decide to answer. This great lack of representation from the population is too big to consider a strong and clear correlation between employers and health care coverage. 

## **Health Care Expedentiture**
For the chart concerning health care expenditure (Figure 2), We can see the proportion of people who spent a specific level of money on health insurance, whether it be zero,less than 500$, or the other levels. For example, we see that the largest proportion of people mostly spent less than 500 dollars through out time, with atleast more than 30% given any year. Specifically, for 2005, 37.5% percent of the population _spent less than 500 dollars_, but for 2015, it went down to 32.04%. For the people _spending no money_ on health insurance, this proportion mostly stayed between 8% to 12% over the years, with a positive increase over time. Meanwhile, The proportion of people _spending between 3000 to 4000 dollars_ doesn't have a linear change over time like the other levels of expedenture. This proportion stayed between 5% and 6% over time. Showing no linear change for _spending between 3000 to 4000 dollars_ can be interpreted as a positive income, since that means the proportion of the population spending money on health insurance are not seeing an increase of expenditure. One would expect that with the introduction of the Affordable Care Act in 2010, which promised reducing health care costs, the amount of money people are spending on health care would decrease. These expectations turned out to be true, since the graph reflects a positive  rate of change for people spending no money on health insurance, and negative rate of change for people spending less than 500 dollars.

# RELATED WORK
Previous projects have been done that measured trends in health insurance coverage over the course of time.  The report was done by the National Center for Health Statistics (NCHS), and presents the selected estimates for health insurance coverage among the civilian noninstitutionalized U.S. population based on data from 2015 National Health Interview Survey (NHIS), as well as comparable estimates from the NHIS 2010-2014 surveys.  From their results, they were able to determine that the trends displayed an overall increase in the population that had health insurance coverage in comparison to previous years.  They then observe short term trends by age, poverty status, as well as race and ethnicity to provide a deeper insight of health insurance coverages, and what might be influencing access to them.

# CONLUSION
Final summary and a description of how your system could be extended.

# REFERENCES

1. IPUMS Health Surveys. Retrieved February 27, 2017
from https://ihis.ipums.org/ihis/index.shtml

2. David Himmelstein,Steffie Woolhandler. January 23, 2017. Repealing the Affordable Care Act will kill more than 43,000 people annually. Retrieved March 6, 2017 from https://www.washingtonpost.com/posteverything/wp/2017/01/23/repealing-the-affordable-care-act-will-kill-more-than-43000-people-annually/?utm_term=.a2fbb24cd075

3.  Robin A. Cohen. May 2016. Health Insurance Coverage: Early Release of
Estimates From the National Health Interview Survey, 2015. Retrieved March 2, 2017 from https://www.cdc.gov/nchs/data/nhis/earlyrelease/insur201605.pdf